\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-15 ace-pldi14/ace-pldi14.tex
pldi14 sectionized sec 3

Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}. When encountering a compound expression (e.g. \verb|e1[e2]|), Ace defers control over typechecking and translation to a \emph{method} associated with the active type of a designated subexpression. The choice of subexpression is fixed by the language, motivated by a consideration of which expression plays a \emph{primary} role (for example, for subscripting, \verb|e1| is chosen). Collectively, these choices comprise the \emph{dispatch protocol} of Ace. Below are examples of these choices. Due to space constraints, we do not show the full protocol, consisting of a rule for each compound form in the grammar \cite{pythonast}.
\begin{itemize}
\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}\todo{is it n-ary?}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g. \verb|-e|) is handed to the type recursively assigned to \texttt{e}.
\item Responsibility over {\bf binary operations} (e.g. \verb|e1 + e2|) is first handed to the type assigned to the left operand. If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call. {Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec. \ref{related} for a discussion.}
\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base).
\end{itemize}

\subsection{Active Bases}\label{abases}
\begin{codelisting}
\lstinputlisting{listing9.py}
\caption{\texttt{[ace.OpenCL]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism, AT\&T.}
\label{pointers}
\end{codelisting}
Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface. The active base specifies the \emph{base semantics} of that function -- the semantics of statements and expressions that do not have a clear ``primary'' subexpression.

@@

\subsection{Monolithic vs. Extensible Languages}
OpenCL is not necessarily the best tool for every job in high-performance computing. Indeed, HPC is an area where designing a set of primitives that satisfy all users has been particularly challenging, and it appears unlikely that a broad consensus will emerge given the variety of architectures, applications, scales and user communities that it serves, and the number of seemingly promising abstractions that emerge continuously targeting various subsets of this problem space.

%Because Ace libraries can contain compile-time logic, written in the metalanguage as described in the previous section, these primitive definitions can be distributed as modules, rather than as extensions to particular compilers or domain-specific languages.
\subsection{Active Typechecking and Translation}
To achieve these criteria, Ace has been designed around a minimal, {extensible} core. Users introduce the compile-time logic associated with primitive types and operations from \emph{within libraries}, as opposed to by some language-external mechanism such as a domain-specific language framework or extensible compiler (see Section \ref{related}).

The two phases of compilation that can be controlled by users are together referred to as \emph{active typechecking and translation (AT\&T)}. They are invoked the first time the \verb|compile| method of a generic function is called, or when type propagation occurs, with a particular type assignment (a cached concrete function is returned subsequently).

\subsubsection{Active Typechecking}

%We now explain how new primitive types and operators can be defined in Ace. 
When the compiler encounters an expression, it must first verify its validity by either assigning it a type or raising a meaningful type error. Rather than defining fixed logic for this, Ace defers control to the {type} of the \emph{primary operand} according to a {\em dispatch protocol}. Because types are metalanguage instances of user-defined types derived from \verb|ace.Type| (cf. Section 2.5), dispatch corresponds to calling a method of this base class.

\subsubsection{Dispatch Protocol}

\subsubsection{User-Defined Backends}\label{backends}
Thus far, we have discussed using OpenCL itself as a backend for our implementation of the OpenCL primitives. This backend is called \verb|OpenCL| in the \verb|ace.OpenCL| module. Ace supports the definition of new backends in a manner similar to the introduction of new types, by extending the \verb|ace.Backend| base class. Backends are specified for a function by using the decorator form \verb|@backend.fn|, as can be seen in the previous examples. 
Backends are responsible for some aspects of the grammar that do not admit simple dispatch to the type of a subterm, such as number and string literals or statement forms like \verb|while| (though \verb|for| is handled by an iterator-like protocol, not shown).

In addition to the OpenCL backend, preliminary C99 and CUDA backends are available (with the caveat that they have not been as fully developed or tested as of this writing.) This allows us to use the OpenCL kernel language, which offers a simplified variant of C99, without relying on the full OpenCL stack, which may not be well-supported by  vendors with proprietary solutions such as CUDA. Backends not based on the C family are also possible, but we leave such developments for future work. During the translation phase, types can access the backend via the context and emit different code for different supported backends (not shown above due to space considerations).

@@

>>>>

Instead of taking a syntax-directed approach, the Ace compiler's typechecking and translation phases take a \emph{type-directed approach}. When encountering a compound expression (e.g. \verb|e1[e2]|), Ace defers control over typechecking and translation to the active type of a designated subexpression. The choice of subexpression is fixed by the language, motivated by considerations of Python's semantics for the same form. Collectively, these choices comprise the \emph{dispatch protocol} of Ace. Below are examples of these choices. Due to space constraints, we do not show the full protocol, consisting of a rule for each compound form in the grammar \cite{pythonast}.
\begin{itemize}
\item Responsibility over {\bf attribute access} (\texttt{e.attr}), {\bf subscripting} (\texttt{e[e1]}\todo{is it n-ary?}) and \textbf{calls} (\verb|e(e1, ..., en)|) and {\bf unary operations} (e.g. \verb|-e|) is handed to the type recursively assigned to \texttt{e}.
\item Responsibility over {\bf binary operations} (e.g. \verb|e1 + e2|) is first handed to the type assigned to the left operand. If it indicates a type error, the type assigned to the right operand is handed responsibility, via a different method call. {Note that this operates like the corresponding rule in Python's \emph{dynamic} operator overloading mechanism; see Sec. \ref{related} for a discussion.}
\item Responsibility over {\bf simple assignment statements}, is handed to the type of the variable on the left (which, as we will see below, is determined by the active base). If this type does not provide special assignment semantics, the base handles it. %Destructuring assignment is also supported by a somewhat more complex protocol.\todo{clarify}
\end{itemize}

\subsubsection{Active Typechecking}
\begin{codelisting}
\lstinputlisting{listing9.py}
\caption{\texttt{[ace.OpenCL]} A portion of the implementation of OpenCL pointer types implementing subscripting logic using the Ace extension mechanism, AT\&T.}
\label{pointers}
\end{codelisting}
When the compiler encounters an expression, it must first verify its validity by either assigning it a type or raising a meaningful type error. Rather than defining fixed logic for this, Ace defers control to the {type} of the \emph{primary operand} according to a {\em dispatch protocol}. Because types are metalanguage instances of user-defined types derived from \verb|ace.Type| (cf. Section 2.5), dispatch corresponds to calling a method of this base class.

\subsection{Active Bases}\label{abases}
Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface. The active base specifies the \emph{base semantics} of that function -- the semantics of statements and expressions that do not have a clear ``primary'' subexpression.

@@

@@

OpenCL is not necessarily the best tool for every job in high-performance computing. Indeed, HPC is an area where designing a set of primitives that satisfy all users has been particularly challenging, and it appears unlikely that a broad consensus will emerge given the variety of architectures, applications, scales and user communities that it serves, and the number of seemingly promising abstractions that emerge continuously targeting various subsets of this problem space.
